---
title: "Stat 4604: Lab 3"
date: "October 19, 2022"
author: "Adam Shen"
format:
  revealjs:
    theme: [dark, "../slides.scss"]
    overview: true
    progress: true
    history: false
    touch: true
    keyboard: true
    controls: true
    controls-back-arrows: visible
    controls-layout: edges
    help: true
    highlight-style: "../adam_one_light.theme"
    from: markdown-smart
    html-math-method: katex
---

```{r}
#| include: false
knitr::opts_chunk$set(
  echo=TRUE, message=FALSE, warning=FALSE, fig.align="center", dev="svglite"
)

old_hooks <- fansi::set_knit_hooks(
  knitr::knit_hooks,
  which = c("output", "message", "warning", "error")
)

options(crayon.enabled=TRUE)
```

## Slide navigation

- Type `?` to bring up the presentation menu

- Press `o` for the presentation overview (and jump to a specific slide)

- To export slides to PDF, press `e` to activate PDF export mode, then print to PDF (note that some
features may not function properly when exported to a PDF)

- To copy a code block, hover over the top-right corner of the code block and click the clipboard
icon

## Today's packages

We will be using the Tidyverse extensively today since we are working with data frames.

```{r, message=TRUE}
library(tidyverse)
```

As usual, packages can be installed using `install.packages("package-name")` and only needs to be
done once per device.

```{r}
theme_set(theme_bw())
```

We will be plotting with the `ggplot2` package later. `theme_set(theme_bw())` sets a plotting theme.

## Read in the data

```{r}
drugs <- read_csv("./data/Pdata.csv")

drugs
```

Note: although the lab instructions say to extract the counts as `N` and the covariates as `X` from
the data set, this is not necessary (and in my opinion, not recommended).

## Fit a basic GLM

```{r}
model <- glm(N ~ ., family="poisson", data=drugs)
```

- In the formula `N ~ .`:

  - The `N` on the left hand side means that we want `N` to be the dependent variable
  
  - The `.` on the right hand side means that we want all other variables appearing in the data set
  (other than `N`) to be used as covariates
  
  - The usage of the `.` in formulas is especially useful when our model uses all variables found
  in a data set, saving us from needing to type out the names of all the variables individually

## Inspect the GLM

```{r}
summary(model)
```

## Non-parametric bootstrap

- The parametric bootstrap is a procedure that begins with simulating the observation of "new" data
by sampling the current data **with replacement**

. . .

- This new resampled data is then used to fit a new model in order to obtain new parameter estimates

. . .

- Once we have resampled our data and fit new models a sufficient number of times, we can aggregate
all the parameter estimates and perform inference

. . .

Task: create a function that performs the parametric bootstrap and returns the parameter estimates
from each re-fitting.

## âœ”ï¸ yourself before you ðŸ’¥ yourself

- We should make sure that our existing and new variable names are clear

  - We will need to compare our bootstrap results with the original model so we don't want to get
  them mixed up
  
  - We want to make sure that we don't end up in a situation where you resample from a recently
  resampled data set

. . .

- Since *one observation* corresponds to *one row* in our data set, we need to make sure that
we are resampling **entire rows** from the data set and that we are **not** resampling the response
values independently of the covariate values

  - In other words, there should not be any "shuffling" of response values among covariate values

## Procedure outline

0. Initialize a variable that will store our parameter estimates. We do not need to store the
resampled data, or the re-fitted models!

1. Create a resample of the original data set.

2. Fit a new model using the resampled data.

3. Extract parameter estimates from the model and store them for later.

4. Repeat steps 1 to 3 a pre-specified number of times.

## Checklist for building our function

Our function should have arguments for:

- The data set that needs to be resampled
  
  - We don't want the function to depend on the existence of a data set in our global environment
  called `drugs`. This will allow us to reuse this function in the future with different data
  sets, if needed
  
. . .
  
- The original model that we created
  
  - We will take advantage of the `update` function that exists in base-R

  - This function will allow us to "swap" the data set used in the original model with our resampled
  data and the model will automatically get re-fitted
  
. . .

- The number of times we wish to repeat the procedure
  
## Build our function {auto-animate=true}

```{r, eval=FALSE}
bootstrap_glm <- function(original_data, original_model, n) {
  out <- matrix(
    nrow=n, ncol=length(coef(original_model)),
    dimnames=list(NULL, names(coef(original_model)))
  )
}
```

We start by creating a matrix with `n` rows, and columns according to the number of parameters in
the original model. This matrix has no row names. The column names will be the names of the
covariates in our original model. We will coerce this matrix to a tibble at the end.

## Build our function {auto-animate=true, visibility=uncounted}

```{r, eval=FALSE}
bootstrap_glm <- function(original_data, original_model, n) {
  out <- matrix(
    nrow=n, ncol=length(coef(original_model)),
    dimnames=list(NULL, names(coef(original_model)))
  )
  
  for (iter in 1:n) {
  
  }
}
```

We will use a `for` loop since we know exactly how many times to repeat the procedure.

## Build our function {auto-animate=true, visibility=uncounted}

```{r, eval=FALSE}
bootstrap_glm <- function(original_data, original_model, n) {
  out <- matrix(
    nrow=n, ncol=length(coef(original_model)),
    dimnames=list(NULL, names(coef(original_model)))
  )
  
  for (iter in 1:n) {
    resampled_data <- original_data %>%
      slice_sample(n=nrow(original_data), replace=TRUE)
  }
}
```

We resample from `original_data` using the `slice_sample` function (From `dplyr`), which returns a
random sample of the rows of a data set. We specify:

- `n=nrow(original_data)` since we want our resampled data to have the same number of rows as the
original data

- `replace=TRUE` because we want to sample with replacement

## Build our function {auto-animate=true, visibility=uncounted}

```{r, eval=FALSE}
bootstrap_glm <- function(original_data, original_model, n) {
  out <- matrix(
    nrow=n, ncol=length(coef(original_model)),
    dimnames=list(NULL, names(coef(original_model)))
  )
  
  for (iter in 1:n) {
    resampled_data <- original_data %>%
      slice_sample(n=nrow(original_data), replace=TRUE)
    
    refitted_model <- original_model %>%
      update(data=resampled_data)
  }
}
```

We pass our original model to the `update` function and specify `data=resampled_data`. This means
that we want to update the fit of our original model by updating the data used to fit the model.

## Build our function {auto-animate=true, visibility=uncounted}

```{r, eval=FALSE}
bootstrap_glm <- function(original_data, original_model, n) {
  out <- matrix(
    nrow=n, ncol=length(coef(original_model)),
    dimnames=list(NULL, names(coef(original_model)))
  )
  
  for (iter in 1:n) {
    resampled_data <- original_data %>%
      slice_sample(n=nrow(original_data), replace=TRUE)
    
    refitted_model <- original_model %>%
      update(data=resampled_data)
    
    out[iter, ] <- coef(refitted_model)
  }
}
```

We update the `iter`th row of the matrix that stores our parameter estimates.

## Build our function {auto-animate=true, visibility=uncounted}

```{r}
bootstrap_glm <- function(original_data, original_model, n) {
  out <- matrix(
    nrow=n, ncol=length(coef(original_model)),
    dimnames=list(NULL, names(coef(original_model)))
  )
  
  for (iter in 1:n) {
    resampled_data <- original_data %>%
      slice_sample(n=nrow(original_data), replace=TRUE)
    
    refitted_model <- original_model %>%
      update(data=resampled_data)
    
    out[iter, ] <- coef(refitted_model)
  }
  
  as_tibble(out)
}
```

We coerce our results into a nice tibble before returning it. Note that this function is general
enough that in addition to `glm` support, it will also supports models of class `lm`.

## Test our function

```{r, cache=TRUE}
set.seed(99)

boot_coefs <- bootstrap_glm(drugs, model, n=10000)
```

This took about 25 seconds to complete. You may wish to decrease `n` if it is taking too long.

. . .

```{r}
boot_coefs
```

## Histogram for bootstrapped $\widehat{\beta}_{1}$

::: {.panel-tabset}

### Code

```{r, eval=FALSE}
ggplot(boot_coefs, aes(x=Trt)) +
  geom_histogram(
    aes(y=after_stat(density)), 
    colour="#56B4E9", fill="#0072B2", alpha=0.7
  )
```

### Plot

```{r, echo=FALSE}
ggplot(boot_coefs, aes(x=Trt)) +
  geom_histogram(
    aes(y=after_stat(density)),
    colour="#56B4E9", fill="#0072B2", alpha=0.7
  )
```

:::

## Overlay a predicted density

::: {.panel-tabset}

### Code

```{r, eval=FALSE}
mean_boot_beta1 <- mean(boot_coefs$Trt)
sd_boot_beta1 <- sd(boot_coefs$Trt)

ggplot(boot_coefs, aes(x=Trt)) +
  geom_histogram(
    aes(y=after_stat(density)),
    colour="#56B4E9", fill="#0072B2", alpha=0.7
  ) +
  geom_function(
    fun=dnorm, args=list(mean=mean_boot_beta1, sd=sd_boot_beta1),
    colour="#D55E00", size=1.2
  )
```

### Plot

```{r, echo=FALSE}
mean_boot_beta1 <- mean(boot_coefs$Trt)
sd_boot_beta1 <- sd(boot_coefs$Trt)

ggplot(boot_coefs, aes(x=Trt)) +
  geom_histogram(
    aes(y=after_stat(density)),
    colour="#56B4E9", fill="#0072B2", alpha=0.7
  ) +
  geom_function(
    fun=dnorm, args=list(mean=mean_boot_beta1, sd=sd_boot_beta1),
    colour="#D55E00", size=1.2
  )
```

:::

## Compute confidence intervals for $\beta_{1}$

Using 95% confidence level.

::: {.panel-tabset}

### Code

```{r, eval=FALSE}
list(
  boot_norm = tibble(
    lwr = mean_boot_beta1 - qnorm(0.975) * sd_boot_beta1,
    upr = mean_boot_beta1 + qnorm(0.975) * sd_boot_beta1
  ),
  
  boot_perc = tibble(
    lwr = quantile(boot_coefs$Trt, 0.025),
    upr = quantile(boot_coefs$Trt, 0.975)
  ),
  
  wald = model %>%
    confint.default() %>%
    as_tibble() %>%
    slice(2) %>%
    rename(lwr=1, upr=2)
)
```

### Output

```{r, echo=FALSE}
list(
  boot_norm = tibble(
    lwr = mean_boot_beta1 - qnorm(0.975) * sd_boot_beta1,
    upr = mean_boot_beta1 + qnorm(0.975) * sd_boot_beta1
  ),
  
  boot_perc = tibble(
    lwr = quantile(boot_coefs$Trt, 0.025),
    upr = quantile(boot_coefs$Trt, 0.975)
  ),
  
  wald = model %>%
    confint.default() %>%
    as_tibble() %>%
    slice(2) %>%
    rename(lwr=1, upr=2)
)
```

:::
